---
title: "Reference Data QC: MegaMUGA"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    code_folding: hide
---

```{r setup, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
library(tidyverse)
library(data.table)
library(purrr)
library(qtl2)
library(magrittr)
library(DT)
library(plotly)
QCtheme <- theme_bw() + 
  theme(panel.grid = element_blank(),
        axis.text = element_text(colour = "black"),
        axis.title = element_text(colour = "black"))
```

# MegaMUGA Annotations

```{r Reading in reference genotypes and metadata}
# Reading in "control genotype" data
megamuga_controls <- data.table::fread("../data/MMControls/MegaMUGA genotypes CONTROLS.csv")
control_genotypes <- data.table::fread("../data/MMControls/control.genotypes.csv")
colnames(control_genotypes)[1] <- "marker"

## Reading in marker annotations
mm_metadata <- data.table::fread("../data/MMControls/mm_uwisc_v2.csv")

## Calculating allele frequencies for each marker
control_allele_freqs <- control_genotypes %>%
  tidyr::pivot_longer(-marker, names_to = "sample", values_to = "genotype") %>%
  dplyr::group_by(marker, genotype) %>%
  dplyr::summarise(n = n()) %>%
  dplyr::mutate(freq = round(n/sum(n), 3),
                genotype = as.factor(genotype)) %>%
  dplyr::left_join(., mm_metadata)
```

First, we searched for probes where many mice are missing genotype calls.

```{r Markers with high "N" counts among reference samples}
## Filtering to markers with missing genotypes
no.calls <- control_allele_freqs %>%
  dplyr::filter(genotype == "N") %>%
  tidyr::pivot_wider(names_from = genotype, values_from = n) %>%
  dplyr::select(marker, chr, bp_grcm39, freq) %>%
  dplyr::mutate(chr = as.factor(chr))
cutoff <- quantile(no.calls$freq, probs = seq(0,1,0.05))[[20]]
above.cutoff <- no.calls %>%
  dplyr::filter(freq > cutoff)
```

Of `r length(unique(as.factor(control_allele_freqs$marker)))` markers, `r nrow(no.calls)` failed to genotype at least one sample, and `r nrow(above.cutoff)` markers failed to genotype at least `r cutoff*100`% of samples.

```{r Plotting no calls, echo=FALSE}

above.cutoff %>%
  dplyr::left_join(.,mm_metadata) %>%
  dplyr::mutate(chr = as.factor(chr)) %>%
  dplyr::arrange(marker) %>%
  DT::datatable(., filte = "top", 
              escape = FALSE, 
              options = list(columnDefs = list(list(width = '20%', targets = c(8)))))

ggplot(no.calls, mapping = aes(x = freq)) + 
  geom_histogram(bins = 100) +
  scale_x_continuous(breaks = seq(0,1,0.1)) + 
  QCtheme + 
  labs(x = "Fraction of mice with missing genotypes",
       y = "Number of markers")
```

In a similar fashion, we calculated the number of reference samples with missing genotypes. Repeated observations of samples/strains meant that genotype counts for each marker among them couldn't be grouped and tallied, so determining no-call frequency occurred column-wise.

```{r Reference samples with high missingness}
n.calls.strains <- apply(X = control_genotypes[,2:ncol(control_genotypes)], MARGIN = 2, function(x) table(x)[5])
n.calls.strains.df <- data.frame(n.calls.strains)
n.calls.strains.df$sample <- names(n.calls.strains)
n.calls.strains.df %<>%
  dplyr::rename(n.no.calls = n.calls.strains)

sampleQC <- ggplot(n.calls.strains.df, 
                   mapping = aes(x = reorder(sample,n.no.calls), 
                                 y = n.no.calls,
                                 text = paste("Sample:", sample))) + 
  geom_point() +
  QCtheme + 
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  labs(x = "Number of mice with missing genotypes",
       y = "Number of markers")
ggplotly(sampleQC, tooltip = c("text","y"))

high.n.samples <- n.calls.strains.df %>%
  dplyr::filter(n.no.calls > quantile(n.calls.strains.df$n.no.calls, probs = seq(0,1,0.05))[20])
```

We next validated the sexes of each sample using Chromosome X probe intensities. We paired up probe intensities, joined available metadata, and filtered down to only markers covering the X chromosome.

```{r Filtering to Chr X Markers}
## Reading in genotype intensities
x_intensities <- data.table::fread("../data/MMControls/control.X.csv")
colnames(x_intensities)[1] <- "marker"
y_intensities <- data.table::fread("../data/MMControls/control.Y.csv")
colnames(y_intensities)[1] <- "marker"
## Check to see if dimensions of intensity tables are identical, marker orders identical, and sample orders identical
if((unique(dim(x_intensities) == dim(y_intensities)) && 
   unique(colnames(x_intensities) == colnames(y_intensities)) &&
   unique(x_intensities$marker == y_intensities$marker)) == TRUE){
     ## Pivoting the data longer
  x_int_long <- x_intensities %>%
  tidyr::pivot_longer(cols = -marker, names_to = "sample", values_to = "x_int")
  y_int_long <- y_intensities %>%
  tidyr::pivot_longer(cols = -marker, names_to = "sample", values_to = "y_int")
  
  long_intensities <- cbind(x_int_long, y_int_long)
} else {
     print("Source intensity data frames have non-identical structure; exiting")
}

## Joining slimmer intensity files with marker metadata and reducing to Chromosome X markers
long_XY_intensities <- long_intensities[,c(1,2,3,6)] %>%
  dplyr::left_join(., mm_metadata) %>%
  dplyr::filter(chr %in% c("X","Y"))

```

Then we flagged markers with high missingness across all samples and samples with high missigness among all markers.

```{r Flagging "low-quality" markers and samples}
## Flagging markers and samples based on previous QC steps
flagged_XY_intensities <- long_XY_intensities %>%
  dplyr::mutate(marker_flag = dplyr::if_else(condition = marker %in% above.cutoff$marker,
                                             true = "FLAG",
                                             false = "")) %>%
  dplyr::mutate(high_missing_sample = dplyr::if_else(condition = sample %in% high.n.samples$sample,
                                                     true = "FLAG",
                                                     false = ""))
```

The first round of inferring predicted sexes used a rough search of the sample name for a proper convention among F1 hybrids, as well as an increasingly large regex search into the sample ID to identify properly named samples from inbred strains.

```{r Preliminary sex prediction}
## First round of predicted sex inference
prelim.predicted.sexes <- flagged_XY_intensities %>%
  dplyr::mutate(bg = dplyr::case_when(stringr::str_detect(string = sample, 
                                                          pattern = "F1") == TRUE ~ "F1",
                                      TRUE ~ as.character(sample)),
                predicted.sex = dplyr::case_when(stringr::str_detect(string = sample, 
                                                          pattern = "F1f") == TRUE ~ "f",
                                      stringr::str_detect(string = sample, 
                                                          pattern = "F1m") == TRUE ~ "m",
                                      TRUE ~ "unknown"))

unknown <- prelim.predicted.sexes %>%
  dplyr::filter(predicted.sex == "unknown")

digit.trim <- unknown %>% 
  dplyr::mutate(mouse.id.1 = stringr::str_sub(sample, -1),
                predicted.sex.1 = dplyr::case_when(stringr::str_sub(mouse.id.1, 1, 1) %in% c("m","M") ~ "m",
                                                   stringr::str_sub(mouse.id.1, 1, 1) %in% c("f","F") ~ "f",
                                                 TRUE ~ "unknown"),
                bg = if_else(condition = (predicted.sex.1 == "m" | predicted.sex.1 == "f"), 
                             true = str_replace(string = bg, 
                                                pattern = mouse.id.1, 
                                                replacement = ""), 
                             false = bg),
                
                mouse.id.3= stringr::str_sub(sample, -3),
                predicted.sex.3 = dplyr::case_when(stringr::str_sub(mouse.id.3, 1, 1) %in% c("m","M") ~ "m",
                                                 stringr::str_sub(mouse.id.3, 1, 1) %in% c("f","F") ~ "f",
                                                 TRUE ~ "unknown"),
                bg = if_else(condition = (predicted.sex.3 == "m" | predicted.sex.3 == "f"), 
                             true = str_replace(string = bg, 
                                                pattern = mouse.id.3, 
                                                replacement = ""), 
                             false = bg),
                
                mouse.id.4 = stringr::str_sub(sample, -4),
                predicted.sex.4 = dplyr::case_when(stringr::str_sub(mouse.id.4, 1, 1) %in% c("m","M") ~ "m",
                                                 stringr::str_sub(mouse.id.4, 1, 1) %in% c("f","F") ~ "f",
                                                 TRUE ~ "unknown"),
                bg = if_else(condition = (predicted.sex.4 == "m" | predicted.sex.4 == "f"), 
                             true = str_replace(string = bg, 
                                                pattern = mouse.id.4, 
                                                replacement = ""), 
                             false = bg),
                
                mouse.id.5 = stringr::str_sub(sample, -5),
                mouse.id.5 = stringr::str_replace(string = mouse.id.5,  ## a couple symbols in these ids mess up the regex search
                                                  pattern = "[:symbol:]", 
                                                  replacement = ""),
                predicted.sex.5 = dplyr::case_when(stringr::str_sub(mouse.id.5, 1, 1) %in% c("m","M") ~ "m",
                                                 stringr::str_sub(mouse.id.5, 1, 1) %in% c("f","F") ~ "f",
                                                 TRUE ~ "unknown"),
                bg = if_else(condition = (predicted.sex.5 == "m" | predicted.sex.5 == "f"),
                             true = str_replace(string = bg,
                                                pattern = mouse.id.5,
                                                replacement = ""),
                             false = bg),
                
                mouse.id.6 = stringr::str_sub(sample, -6),
                mouse.id.6 = stringr::str_replace(string = mouse.id.6,  ## a couple symbols in these ids mess up the regex search
                                                  pattern = "[:punct:]", 
                                                  replacement = ""),
                mouse.id.6 = stringr::str_replace(string = mouse.id.6,  ## a couple symbols in these ids mess up the regex search
                                                  pattern = "[:symbol:]", 
                                                  replacement = ""),
                predicted.sex.6 = dplyr::case_when(stringr::str_sub(mouse.id.6, 1, 1) %in% c("m","M") ~ "m",
                                                 stringr::str_sub(mouse.id.6, 1, 1) %in% c("f","F") ~ "f",
                                                 TRUE ~ "unknown"),
                bg = if_else(condition = (predicted.sex.6 == "m" | predicted.sex.6 == "f"), 
                             true = str_replace(string = bg, 
                                                pattern = mouse.id.6, 
                                                replacement = ""), 
                             false = bg)) %>%
  dplyr::mutate(predicted.sex = dplyr::case_when(predicted.sex.1 == "m" ~ "m", 
                                                 predicted.sex.3 == "m" ~ "m",
                                                 predicted.sex.4 == "m" ~ "m",
                                                 predicted.sex.5 == "m" ~ "m",
                                                 predicted.sex.6 == "m" ~ "m",
                                                 
                                                 predicted.sex.1 == "f" ~ "f", 
                                                 predicted.sex.3 == "f" ~ "f",
                                                 predicted.sex.4 == "f" ~ "f",
                                                 predicted.sex.5 == "f" ~ "f",
                                                 predicted.sex.6 == "f" ~ "f",
                                                 TRUE ~ "unknown"))

predicted.sexes.strings <- prelim.predicted.sexes %>%
  dplyr::filter(predicted.sex != "unknown") %>%
  dplyr::bind_rows(.,digit.trim)

## Taking the first marker as a sample
predicted.sex.table <- predicted.sexes.strings %>%
  dplyr::filter(marker %in% unique(prelim.predicted.sexes$marker)[1]) %>%
  dplyr::select(sample, predicted.sex, bg) %>%
  dplyr::group_by(predicted.sex) %>%
  dplyr::count() 

```

This captured `r predicted.sex.table[which(predicted.sex.table$predicted.sex == "f"),2]$n` female samples, `r predicted.sex.table[which(predicted.sex.table$predicted.sex == "m"),2]$n` male samples, leaving `r predicted.sex.table[which(predicted.sex.table$predicted.sex == "unknown"),2]$n` samples of unknown predicted sex from nomenclature alone. These samples exhibit a range of naming inconsistencies.

```{r}
predicted.sexes.strings %>%
  dplyr::filter(predicted.sex == "unknown",
                marker == predicted.sexes.strings$marker[[1]]) %>%
  dplyr::select(sample, bg)
```

After predicting the sexes of the vast majority of reference samples, we visualized the average probe intensity among X Chromosome markers for each sample, labeling them by predicted sex.

```{r}
mean.x.intensities.by.sex <- predicted.sexes.strings %>%
  dplyr::filter(marker_flag != "FLAG") %>%
  dplyr::group_by(sample, predicted.sex, chr, high_missing_sample) %>%
  dplyr::summarise(mean.x = mean(x_int),
                   mean.y = mean(y_int))

predicted.sex.plot.palettes <- mean.x.intensities.by.sex %>%
  dplyr::ungroup() %>%
  dplyr::distinct(sample, predicted.sex, high_missing_sample) %>%
  dplyr::mutate(predicted.sex.palette = dplyr::case_when(predicted.sex == "f" ~ "#5856b7",
                                                         predicted.sex == "m" ~ "#eeb868",
                                                         predicted.sex == "unknown" ~ "black"))

predicted.sex.palette <- predicted.sex.plot.palettes$predicted.sex.palette
names(predicted.sex.palette) <- predicted.sex.plot.palettes$predicted.sex

mean.x.intensities.by.sex.plot <- ggplot(mean.x.intensities.by.sex, 
                                         mapping = aes(x = mean.x, 
                                                       y = mean.y, 
                                                       colour = predicted.sex,
                                                       shape = high_missing_sample,
                                                       text = sample,
                                                       label = high_missing_sample)) + 
  geom_point() + 
  scale_colour_manual(values = predicted.sex.palette) + 
  facet_grid(.~chr) +
  QCtheme
ggplotly(mean.x.intensities.by.sex.plot, 
         tooltip = c("text","label"),)
```